Type Classes for Mathematics in Type Theory

### 2010-11-26 papers/mscs/mscs.tex
First batch of MSCS paper modifications addressing concerns brought up by referees.

More elaborate structures, too, can be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in principle, in practice \emph{na\"ive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.

@@

In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and \textsc{Ssreflect}~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: ∀ x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a canonical name for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.

@@

We will tackle the problems associated with the use of structure predicates one by one, starting with those encountered during theorem \emph{application}. Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows\footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate. The record form is simply more convenient in that it immediately gives us named projections for laws and substructures.}:
\begin{lstlisting}
  Record SemiGroup (G: Type) (e: relation G) (op: G → G → G): Prop :=
    { sg_setoid: Equivalence e
    ; sg_ass: Associative op
    ; sg_proper: Proper (e ==> e ==> e) op }.
\end{lstlisting}
Then by (1) making \lstinline|SemiGroup| a \emph{class} (by replacing the \lstinline|Record| keyword with the \lstinline|Class| keyword), (2) marking its proofs as \emph{instances} (again by replacing a single keyword), and (3) marking the \lstinline|SemiGroup| parameter of semigroup theorems as implicit (by using curly instead of round brackets), we no longer have to pass \lstinline|SemiGroup| proofs around manually ourselves, letting instance resolution do it for us instead. Because instance resolution is part of the unifier, this also works when the statement of the theorem we wish to apply only mentions some of the components (which admittedly doesn't make much sense for semigroups).

@@

Because predicate classes only provide contextual information and are insulated from the actual algebraic expressions, their instances can always be kept entirely opaque---only their existence matters. Together, these properties largely defuse an argument occasionally voiced against type classes concerning perceived unpredictability of instance resolution. While it is certainly true that in contexts with redundant information it can become hard to predict which instance of a predicate class will be found by proof search, it simply \emph{does not matter} which one is found. Moreover, for operational type classes the issue rarely arises because their instances are not nearly as abundant.

@@

We use names for properties like distributivity and absorption, because these are type classes as well (which is why we declare them with \lstinline|:>|). It has been our experience that almost any generic predicate worth naming is worth representing as a predicate type class, so that its proofs will be resolved as instances behind the scenes whenever possible. Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular. Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a `database' of properties previously proved. Much more so than the existing canonical structures facility, because canonical structures can only be used with bundled representations, which we argued against in section~\ref{bundling}.

@@

The \lstinline|monmor_to| and \lstinline|monmor_from| fields are not an absolute necessity, but eliminate the need for theorems to declare \lstinline|Monoid| parameters when they already declare \lstinline|Monoid_Morphism| parameters. This is an instance where we can freely posit structural properties without worrying about potential problems when such information turns out to be redundant in contexts where the source and target of the morphism are already known to be \lstinline|Monoid|s.

@@

A multisorted signature enumerates sorts, operations, and specifies the `types' of the operations as non-empty lists of sorts, where the final element denotes the result type:
\begin{lstlisting}
  Inductive Signature: Type :=
    { sorts: Set
    ; operation:> Set
    ; operation_type:> operation → ne_list sorts }.
\end{lstlisting}
Given an interpretation of the sorts (mapping each symbolic sort to a carrier), an interpretation for the operations is easily written as an operational type class:
\begin{lstlisting}
  Variables (σ: Signature) (carriers: sorts σ → Type).

  Definition op_type: ne_list (sorts σ) → Type := fold (→) ∘ map carrier.

  Class AlgebraOps (σ: Signature) (carriers: sorts σ$$ → Type)
    := algebra_op: ∀ o: operation σ, op_type carriers (σ o).
\end{lstlisting}
Because our carriers will normally be equipped with a setoid equality, we further define the predicate class \lstinline|Algebra|, stating that each of the operations respects the setoid equality on the carriers:
\begin{lstlisting}
  Class Algebra `{∀$$a, Equiv (carriers a)} `{AlgebraOps σ$$ carriers}: Prop :=
      { algebra_setoids:> ∀ a, Setoid (carriers a)
      ; algebra_propers:> ∀ o: σ, Proper (=) (algebra_op o) }.
\end{lstlisting}
The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an \lstinline|Equiv| instance expressing setoid-respecting extensionality for the function types generated by \lstinline|op_type|.

@@

To characterize structures such as semirings and rings, we need equational theories, consisting of a signature together with laws (represented by a predicate over equality entailments):
\begin{lstlisting}
  Record EquationalTheory :=
    { eqt_sig:> Signature
    ; eqt_laws:> EqEntailment eqt_sig → Prop }.
\end{lstlisting}
An \lstinline|EqEntailment| consists of premises and a conclusion represented by an inductively defined statement grammar, which in turn uses an inductively defined term grammar. A detailed discussion of these definitions and the theory developed for them is beyond the scope of this paper.

@@

To give a further taste of what universal algebra in our development looks like, we consider the definitions involved in the first homomorphism theorem~\cite{meinke1993universal} in more detail:
\begin{theorem}[First homomorphism theorem]
If $A$ and $B$ are algebras, and $f$ is a homomorphism from $A$ to $B$, then the equivalence relation defined by $a\sim b$ if and only if $f(a)=f(b)$ is a congruence on $A$, and the algebra $A/\hspace{-1mm}\sim$ is isomorphic to the image of $f$, which is a subalgebra of B.
\end{theorem}

@@

For the homomorphism theorem, we begin by declaring our dramatis personae:
\begin{lstlisting}
  Context `{Algebra σ$$ A} `{Algebra σ$$ B} `{HomoMorphism σ$$ A B f}.
\end{lstlisting}
With \lstinline|~| defined as described, the first part of the proof is simply the definition of the following instance:
\begin{lstlisting}
  Instance co: Congruence σ$$ (~).
\end{lstlisting}

@@

However, there are contexts (in other proofs) where \lstinline|Prop|-sorted construction of subalgebras really \emph{is} appropriate. Unfortunately, Coq's universe polymorphism is not yet up to the task of letting us use a single set of definitions to handle both cases. In particular, there is no universe polymorphism for definitions (as opposed to inductive definitions) yet. We will return to this point later. In our development, we have two sets of definitions, one for \lstinline|Prop| and one for \lstinline|Type|, resulting in duplication of about a hundred lines of code.

>>>>

More elaborate structures, too, can be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in principle, in practice \emph{naive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.

@@

In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and \textsc{Ssreflect}~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\newpage
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: ∀ x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a canonical name for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.

@@

To show that the fully unbundled approach with structures represented by predicates can be made feasible using type classes, we will tackle one by one the problems traditionally associated with their use, starting with those encountered during theorem \emph{application}. Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows\footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate. The record form is simply more convenient in that it immediately gives us named projections for laws and substructures.}:
\begin{lstlisting}
  Record SemiGroup (G: Type) (e: relation G) (op: G → G → G): Prop :=
    { sg_setoid: Equivalence e
    ; sg_ass: Associative op
    ; sg_proper: Proper (e ==> e ==> e) op }.
\end{lstlisting}
Then by (1) making \lstinline|SemiGroup| a \emph{class} (by replacing the \lstinline|Record| keyword with the \lstinline|Class| keyword), (2) marking its proofs as \emph{instances} (again by replacing a single keyword), and (3) marking the \lstinline|SemiGroup| parameter of semigroup theorems as implicit (by using curly instead of round brackets), we no longer have to pass \lstinline|SemiGroup| proofs around manually ourselves, letting instance resolution do it for us instead. Because instance resolution is part of the unifier, this also works when the statement of the theorem we wish to apply only mentions some of the components (which admittedly doesn't make much sense for semigroups).

@@

Because predicate classes only provide contextual information and are insulated from the actual algebraic expressions, their instances can always be kept entirely opaque---only their existence matters. Together, these properties largely defuse an argument occasionally voiced against type classes concerning perceived unpredictability of instance resolution. While it is certainly true that in contexts with redundant information it can become hard to predict which instance of a predicate class will be found by proof search, it simply \emph{does not matter} which one is found. Moreover, for operational type classes the issue rarely arises because their instances are not nearly as abundant, and are systematically shared.

@@

We use names for properties like distributivity and absorption, because these are type classes as well (which is why we declare their instances with \lstinline|:>|). It has been our experience that almost any generic predicate worth naming is worth representing as a predicate type class, so that its proofs will be resolved as instances behind the scenes whenever possible. Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular. Indeed, we believe that type classes provide an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a `database' of properties previously proved.

@@

We include the \lstinline|monmor_to| and \lstinline|monmor_from| fields because it does not make much sense to talk about monoid morphisms between non-monoids, and having these fields removes the need for \lstinline|Monoid| class constraints when we are already parameterizing definitions or theory on a \lstinline|Monoid_Morphism|. On the other hand, we will also wish to talk about monoid morphisms between \emph{known} monoids, and in these cases the fields will be strictly redundant. As mentioned before, it is a strength of our approach that such redundant knowledge is entirely harmless, so that we may freely posit these structural properties whenever they make sense and provide convenience, without risking rebundling tar-pits or projection path ambiguities down the line.

@@

A multisorted signature enumerates sorts, operations, and specifies the `types' of the operations as non-empty lists of sorts, where the final element denotes the result type:
\begin{lstlisting}
  Inductive Signature: Type :=
    { sorts: Set
    ; operation:> Set
    ; operation_type:> operation → ne_list sorts }.
\end{lstlisting}
Given an interpretation of the sorts (mapping each symbolic sort to a carrier type), interpretations of the operations are easily represented by an operational type class:
\begin{lstlisting}
  Variables (σ: Signature) (carriers: sorts σ$$ → Type).

  Class AlgebraOps :=
    algebra_op: ∀ o: operation σ, fold (→$\hspace{-1mm}$) (map carriers (operation_type σ$$ o)).
\end{lstlisting}
Because our carriers will normally be equipped with a setoid equality, we further define the predicate class \lstinline|Algebra|, stating that each of the operations respects the setoid equality on the carriers:
\begin{lstlisting}
  Class Algebra `{∀$$a, Equiv (carriers a)} `{AlgebraOps}: Prop :=
      { algebra_setoids:> ∀ a, Setoid (carriers a)
      ; algebra_propers:> ∀ o: σ, Proper (=) (algebra_op o) }.
\end{lstlisting}
The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an automatically derived \lstinline|Equiv| instance expressing setoid-respecting extensionality for the function types produced by the \lstinline|fold| in \lstinline|AlgebraOps|.

@@

To adequately characterize structures such as semirings and rings, we need not just a signature that enumerates and gives the types of their operations, but also a specification of the axioms (laws) that these operations must satisfy. For this, we define \lstinline|EquationalTheory| as a signature together with a set of laws, the latter represented by a predicate over equality entailments:
\begin{lstlisting}
  Record EquationalTheory :=
    { eqt_sig:> Signature
    ; eqt_laws:> EqEntailment eqt_sig → Prop }.
\end{lstlisting}
An \lstinline|EqEntailment| consists of premises and a conclusion represented by an inductively defined statement grammar, which in turn uses an inductively defined term grammar. A detailed discussion of these definitions and the theory developed for them is beyond the scope of this paper.

@@

To give a further taste of what universal algebra in our development looks like, we consider the definitions involved in the first homomorphism theorem~\cite{meinke1993universal} in more detail:
\begin{theorem}[First homomorphism theorem]
If $A$ and $B$ are algebras, and $f$ is a homomorphism from $A$ to $B$, then the equivalence relation $\sim$ defined by ``$a\sim b \leftrightarrow f(a)=f(b)$'' is a congruence on $A$, and the quotient algebra $A/\hspace{-1mm}\sim$ is isomorphic to the image of $f$, which is a subalgebra of B.
\end{theorem}

@@

For the homomorphism theorem, we begin by declaring our dramatis personae:
\begin{lstlisting}
  Context `{HomoMorphism σ$$ A B f}.
\end{lstlisting}
With \lstinline|~| defined as indicated, the first part of the proof is simply the definition of the following instance:
\begin{lstlisting}
  Instance co: Congruence σ$$ (~).
\end{lstlisting}

@@

However, there are contexts (in other proofs) where \lstinline|Prop|-sorted construction of subalgebras really \emph{is} appropriate. Unfortunately, Coq's universe polymorphism is not yet up to the task of letting us use a single set of definitions to handle both cases. In particular, there is no universe polymorphism for ordinary definitions (as opposed to inductive definitions) yet. We will return to this point later. In our development, we have two sets of definitions, one for \lstinline|Prop| and one for \lstinline|Type|, resulting in duplication of about a hundred lines of code.
